---
title: "Data Analysis - NYPD Shooting"
date: "04/23/2023"
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Dataset information
* Observations: 25.6K, a shooting incident for each
* Variables: 19
* List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

    __Attributes__

    * INCIDENT_KEY : Randomly generated persistent ID for each arrest (Plain Text)
    * OCCUR_DATE : Exact date of the shooting incident (Date & Time)
    * OCCUR_TIME : Exact time of the shooting incident (Plain Text)
    * BORO : Borough where the shooting incident occurred (Plain Text)
    * PRECINCT : Precinct where the shooting incident occurred (Number)
    * JURISDICTION_CODE : Jurisdiction where the shooting incident occurred. Jurisdiction codes 0(Patrol), 1(Transit) and 2(Housing) represent NYPD whilst codes 3 and more represent non NYPD jurisdictions (Number)
    * LOCATION_DESC : Location of the shooting incident (Plain Text)
    * STATISTICAL_MURDER_FLAG : Shooting resulted in the victim's death which would be counted as a murder (Bool)
    * PERP_AGE_GROUP : Perpetrator's age within a category (Plain Text)
    * PERP_SEX : Perpetrator's sex description (Plain Text)
    * PERP_RACE : Perpetrator's race description (Plain Text)
    * VIC_AGE_GROUP : Victim's age within a category (Plain Text)
    * VIC_SEX : Victim's sex description (Plain Text)
    * VIC_RACE : Victim's race description (Plain Text)
    * X_COORD_CD : Midblock X-coordinate for New York State Plane Coordinate System, Long Island Zone, NAD 83, units feet (FIPS 3104) (Plain Text)
    * Y_COORD_CD : Midblock Y-coordinate for New York State Plane Coordinate System, Long Island Zone, NAD 83, units feet (FIPS 3104) (Plain Text)
    * Latitude : Latitude coordinate for Global Coordinate System, WGS 1984, decimal degrees (EPSG 4326) (Number)
    * Longitude : Longitude coordinate for Global Coordinate System, WGS 1984, decimal degrees (EPSG 4326) (Number)
    
[<u>Reference</u>](https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8 )

#### ETL Process

```{r pkg, message=FALSE}
# Packages used in the project
require(tidyverse)
require(lubridate)
require(cowplot) # Used to align ggplot
require(leaflet) # Mark incidents on a map
library(jtools) # Print summary of the model
```


Load data and see what's the data look like.


```{r df}
# Read data
df <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
# Preview
head(df)
```

I usually change all the column names to lower case, and we can do this via the function, `tolower()`.  
Then, select the columns we need in the analysis.

```{r change_colname}
# Transform the column names to lowercase
names(df) <- tolower(names(df))

# The following columns would be used for analysis
df <- df %>% 
    select(
        c(occur_date, occur_time, location_desc,
          statistical_murder_flag, perp_age_group, perp_sex, perp_race,
          vic_age_group, vic_sex, vic_race, latitude, longitude)
    )

# Preview
head(df)
```


From the result, we can see the data type of date variable is not a date object, 
so we can change it via the function, `as.Date()`. 
Then, I'd like to add a few columns related to date and time in order to view the dataset in different time ranges. 
By the way, **though hour, month and year are numbers, I think it's better to see them as categorical variables rather than continuous ones**.
It can be done through the function, `as.factor()`.

```{r add_cols_and_change_dtype}
# Use date data type for occur_date
df$occur_date <- as.Date(df$occur_date, format = "%m/%d/%Y")

# Get the hour & month when the incidents happened
df <- df %>%
    mutate(hour_of_the_day = hour(occur_time)) %>%
    mutate(month = month(occur_date)) %>% 
    mutate(year = year(occur_date))

# Change data type
df$hour_of_the_day <- as.factor(df$hour_of_the_day)
df$month <- as.factor(df$month)
df$year <- as.factor(df$year)

# Preview
head(df)
```


Okay, it seems we have done the basic data cleaning. Yeah~ 🎉 

Let's take a look of the summary. 


```{r summ}
summary(df)
```


Oops, it occurs to me that I forgot an important step, handling missing values.  
Let's see how many `NAs` among each column.


```{r na_values}
# Number of NAs in each columns
print(colSums(is.na(df)))
```

Generally, missing values are among two different columns related to **location** and **perpetrator**.


- For `location_desc`, since its data type is text, it can be excluded from the analysis.

- For `perp_age_group`, `perp_sex`, and `perp_race`, they are important variables and I will still use them in the analysis.
However, there are too many missing values among them, so it's hard to find a suitable method to fill the NAs.
Therfore, dropping the observations which containing NAs might be a more appropriate choice.

```{r}
# Exclude location_desc
df <- df %>% 
    select(-c(location_desc))

# Remove rows with NA's 
df <- drop_na(df)
```
```{r}
print(colSums(is.na(df)))
```

Great! We have done the ETL process. Next, we will start **EDA(Exploratory Data Analysis)** process through visualization.
It's my favorite part of the analysis, let's go!

#### EDA 

The data was recorded for 15 years, we can group the cases in different time span, and see how the cases change.

```{r fig1, fig.width=18, fig.height=6}
# Distribution of the incidents by hour_of_the_day, month, year

# By hour_of_the_day
p1 <- df %>% ggplot(aes(x = hour_of_the_day)) +
    geom_bar()+
    labs(title = "Distribution Of The Incidents by Hour")+
    xlab("Hour Of The Day")+
    ylab("Incidents")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# By month
p2 <- df %>% ggplot(aes(x = month))+
    geom_bar()+
    labs(title = "Distribution Of The Incidents by Month")+
    xlab("Month")+
    ylab("")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# By year
p3 <- df %>% ggplot(aes(x = year))+
    geom_bar()+
    labs(title = "Distribution Of The Incidents by Year")+
    xlab("Year")+
    ylab("")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# Put the plots together
plot_grid(p1, p2, p3, align = "h", nrow = 1)
```

By observing these three plots, 

1. we can see the shooting incidents tend to happen at night, and from May to September, the incidents are much more than the other months.

2. The incidents seem to decline year by year; however, the incidents soared in 2020 and 2021.
The reason behind the fact is worth finding.

```{r fig2, fig.width=12, fig.height=6}
# Distribution of the incidents by vic_age_group, vic_sex, &  perp_age_group, perp_sex

# By vic_age_group, vic_sex
p1 <- df %>% ggplot(aes(x = vic_age_group, fill = vic_sex))+
    geom_bar()+
    labs(title = "Victim's Age Group By Sex")+
    xlab("Age Group")+
    ylab("Incidents")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# By perp_age_group, perp_sex
p2 <- df %>% ggplot(aes(x = perp_age_group, fill = perp_sex))+
    geom_bar() +
    labs(title = "Perpetrator's Age Group By Sex")+
    xlab("Age Group")+
    ylab("")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# Put the plots together
plot_grid(p1, p2)
```

By observing these two plots, 

1. we can see that most victoms and perpetrators are male.


```{r fig3, fig.width=18, fig.height=8}
# Distribution of the incidents by vic_age_group, vic_race, &  perp_age_group, perp_race

# By vic_age_group, vic_race
p1 <- df %>% ggplot(aes(x = vic_age_group, fill = vic_race))+
    geom_bar()+
    labs(title = "Victim's Age Group By Race")+
    xlab("Age Group")+
    ylab("Cases")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# By perp_age_group, perp_race
p2 <- df %>% ggplot(aes(x = perp_age_group, fill = perp_race))+
    geom_bar() +
    labs(title = "Perpetrator's Age Group By Race")+
    xlab("Age Group")+
    ylab("")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

# Put the plots together
plot_grid(p1, p2)
```

By observing these two plots, 

1. we can see that most the race of victoms and perpetrators are black.

```{r fig4, fig.width=5, fig.height=5}
# Whether The Victim Was Murdered

p1 <- df %>% ggplot(aes(x = vic_age_group, fill = statistical_murder_flag))+
    geom_bar()+
    labs(title = "Whether The Victim Was Murdered ", fill = "Murder")+
    xlab("Age Group")+
    ylab("Cases")+
    theme(
    plot.title = element_text(color="steelblue", size=14, face="bold.italic"),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
    )

p1
```
From the data, we can see that most of the victoms survived during the incidents. However, I think it would be better if we can know more about whether the injury was serious or not.

```{r map}
# Mark the incidents on a map
map <- leaflet(df) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addCircles(lat = ~latitude, lng = ~longitude)
map
```

#### Bias

- Personal Bias

    Since I am not an American and I have never been to the US, my understanding of America is
    mostly from news and it's limited. If my language offended you, that was not my intention. I apologize.

    Thus, before I make plots related to sex and race, I thought the victoms would tend to be female. 
    On the other hand, I thought the perpetrators would mostly be white and the victoms might mostly be black.

    However, after making these plots, and examine the results, my bias is mitigated. 

- Data Bias

    From my observation, I could not find detailed information about how the data was recorded and who recorded them. Besides, the publisher didn't state that if there were any measures implemented to avoid bias.
    
    Therefore, the data may contain personal bias if the people who recorded it weren't aware of their own bias. In my opinion, I think it would be better for the publisher to give more detailed information about how they prevented personal bias.




#### Modeling

I implemented Linear Regression to see the relationship between `statistical_murder_flag` and `vic_age_group`,
with different category encoding techniques applied to the two columns.

```{r modeling}
# Use Linear Regression Model

# Transform the target as 0(False) and 1(True)
df$statistical_murder_flag <- as.integer(as.logical(df$statistical_murder_flag))

# Category encoding function, use leave one out technique
encode_leave_one_out <- function(x, y, sigma = NULL) {
    n <- length(x)
    x[is.na(x)] <- "__MISSING"
    x2 <- vapply(1:n, function(i) {
      xval <- x[i]
      yloo <- y[-i]
      xloo <- x[-i]
      yloo <- yloo[xloo == xval]
      mean(yloo, na.rm = TRUE)
    }, numeric(1))
    if (!is.null(sigma)) {
      x2 <- x2 * rnorm(n, mean = 1, sd = sigma)
    }
    x2
}

# Category encoding
df2 <- df
set.seed(18)
df2[["vic_age_group_encoded"]] <- encode_leave_one_out(df[["vic_age_group"]], df[["statistical_murder_flag"]], sigma = 0.05)
head(df2)

# Modeling
mod <- lm(statistical_murder_flag ~ vic_age_group_encoded, data = df2)
# Check the summary of the model
summ(mod)
```

From the model summary, the p-value is less than the significance level(0.05), so we can conclude that the victoms' age is correlated with the target, `statistical_murder_flag`.

